PhD Thesis

### 2017-01-18 aosd13/Evaluation.tex
Rearrange part I chapters
Revise sectioning
s/subsection/section for sections that became chapters of part I.

\newcommand{\graphPath}[1]{aosd13/graphs/#1}
\input{aosd13/graphs/evalRes}
\section{Evaluation}
\label{sec:evaluation}

@@

\subsection{Study Setup}
%evaluation strategy
Throughout this chapter, we have already shown several compact queries for which our optimizations increase performance significantly compared to a naive execution. Since some optimizations change
the complexity class of the query (e.g.\ by using an index), so the speedups grow with the size of the data. However, to get a more realistic evaluation of \LoS{}, we decided to perform an experiment with existing real-world queries.

@@

\subsection{Experimental Units}

@@

\subsection{Measurement Setup}
To measure performance, we executed the queries on the preinstalled JDK class library (\texttt{rt.jar}), containing 58M of uncompressed Java bytecode.
We also performed a preliminary evaluation by running queries on the much smaller ScalaTest library, getting comparable results that we hence do not discuss.
Experiments were run on a 8-core Intel Core i7-2600, 3.40 GHz, with 8 GB of RAM, running Scientific Linux release 6.2.
The benchmark code itself is single-threaded, so it uses only one core; however the JVM used also other cores to offload garbage collection.
We used the preinstalled OpenJDK Java version 1.7.0\_05-icedtea and Scala 2.10.0-M7.

@@

\subsection{Results}

>>>>

\newcommand{\graphPath}[1]{aosd13/graphs/#1}
\input{aosd13/graphs/evalRes}
\chapter{Evaluating SQuOpt}
\label{sec:evaluation}

@@

\section{Study Setup}
%evaluation strategy
Throughout this chapter, we have already shown several compact queries for which our optimizations increase performance significantly compared to a naive execution. Since some optimizations change
the complexity class of the query (e.g.\ by using an index), so the speedups grow with the size of the data. However, to get a more realistic evaluation of \LoS{}, we decided to perform an experiment with existing real-world queries.

@@

\section{Experimental Units}

@@

\section{Measurement Setup}
To measure performance, we executed the queries on the preinstalled JDK class library (\texttt{rt.jar}), containing 58M of uncompressed Java bytecode.
We also performed a preliminary evaluation by running queries on the much smaller ScalaTest library, getting comparable results that we hence do not discuss.
Experiments were run on a 8-core Intel Core i7-2600, 3.40 GHz, with 8 GB of RAM, running Scientific Linux release 6.2.
The benchmark code itself is single-threaded, so it uses only one core; however the JVM used also other cores to offload garbage collection.
We used the preinstalled OpenJDK Java version 1.7.0\_05-icedtea and Scala 2.10.0-M7.

@@

\section{Results}

### 2017-01-18 aosd13/paperBody.tex
Revise sectioning
s/subsection/section for sections that became chapters of part I.

\subsection{Adapting a Query}
\label{subsec:adaptingaquery}
\begin{figure}
\centering
\begin{lstlisting}
import squopt._
import schema.squopt._

@@

\subsection{Indexing}
%We reuse path indexes~\citep{Bertino89}.
%\pg{We have no example which shows a hierarchical index, nor do we say we use
%path indexes, nor do we cite the paper (citation key Bertino89). Is that OK?}
% PG: yes at least for now, because of space reasons.

@@

\subsection{Expression Trees}
\label{sec:representation}
In order to analyze and optimize collection queries at runtime, {\LoS} reifies their syntactic structure as \emph{expression trees}.
The expression tree reflects the syntax of the query after desugaring, that is, after for-comprehensions have been replaced by API calls.
For instance, \code{recordsQuery} from Fig.~\ref{fig:reifiedQuery} points to the
following expression tree (with some boilerplate omitted for clarity):

@@

\subsection{Optimizations}

@@

\subsection{Query Execution}
\label{sec:execution}
Calling the \code{eval} method on a query will convert it to executable
bytecode; this bytecode will be loaded and invoked by using Java reflection.
We produce a thunk that, when evaluated, will execute the
generated code.

@@

\subsection{Implementation: Expressing the interface in Scala}
\label{sec:intfScala}
\input{aosd13/pepm13/caseStudyImpl}

@@

\smartParagraph{Language-Integrated Queries}
Microsoft's Language-Integrated Query technology (\LINQ)~\citep{Meijer:2006:LRO:1142473.1142552,Bierman:2007:LTF:1297027.1297063} is similar to our work in that it also
reifies queries on collections to enable analysis and optimization. Such queries can be executed against a variety of backends (such as SQL databases or in-memory objects), and adding new back-ends is supported. Its implementation uses \emph{expression trees}, a compiler-supported
implicit conversion between expressions and their reification as a syntax tree. There are various major differences, though.
First, the support for expression trees is hard-coded into the compiler. This means that the techniques are not applicable in languages 
that do not explicitly support expression trees. More importantly, the way expression trees are created in \LINQ\ is generic and fixed.
For instance, it is not possible to create different tree nodes for method calls that are relevant to an analysis (such as the \code{map} method) than for method calls that are irrelevant for the analysis (such as the \code{toString} method). For this reason, expression trees in \LINQ\ 
cannot be customized to the task at hand and contain too much low-level information. It is well-known that this makes it quite hard to
implement programs operating on expression trees~\citep{Eini11Pain}.

@@

\smartParagraph{Query Optimization}
Query optimization on relational data is a long-standing issue in the database community, but there
are also many works on query optimization on objects~\citep{Fegaras00,Grust99PhD}.
Compared to these works, we have only implemented a few simple query optimizations, so there is potential
for further improvement of our work by incorporating more advanced optimizations.

@@

\smartParagraph{Scala and DSL Embedding}
\label{sec:rwdsl}
Technically, our implementation of \LoS\ is a deep embedding of a part of the Scala collections API~\citep{odersky2009fighting}.
Deep embeddings were pionereed by \citet{Leijen99DSEC} and \citet{elliott03compiling}. \pg{This makes no sense for the thesis.}
The technical
details of the embedding are not the main topic of this chapter; we are using some of the
Scala techniques presented by \citet{rompf2010lightweight} for using implicits and for
adding infix operators to a type. Similar to \citet{rompf2010lightweight}, we also 
use the Scala compiler on-the-fly. A plausible alternative backend for \LoS\ would
have been to use Delite~\citep{Rompf11BBlocks}, a framework for
building highly efficient DSLs in Scala.
Using this framework, in concurrent work, \citet{Rompf13} also optimize
collection queries; while their work allows for imperative programs, they do
not support embedding arbitrary libraries in an automated way. On the other
hand, they can reuse support for automatic parallelization and multiple platforms present in Delite.
\citet{Ackermann12} present Jet, which also optimizes collection queries but
targets MapReduce-style computations in a distributed environment.
Moreover, both works do not apply typical database optimizations such as
indexing or filter hoisting.

>>>>

\section{Adapting a Query}
\label{subsec:adaptingaquery}
\begin{figure}
\centering
\begin{lstlisting}
import squopt._
import schema.squopt._

@@

\section{Indexing}
%We reuse path indexes~\citep{Bertino89}.
%\pg{We have no example which shows a hierarchical index, nor do we say we use
%path indexes, nor do we cite the paper (citation key Bertino89). Is that OK?}
% PG: yes at least for now, because of space reasons.

@@

\section{Expression Trees}
\label{sec:representation}
In order to analyze and optimize collection queries at runtime, {\LoS} reifies their syntactic structure as \emph{expression trees}.
The expression tree reflects the syntax of the query after desugaring, that is, after for-comprehensions have been replaced by API calls.
For instance, \code{recordsQuery} from Fig.~\ref{fig:reifiedQuery} points to the
following expression tree (with some boilerplate omitted for clarity):

@@

\section{Optimizations}

@@

\section{Query Execution}
\label{sec:execution}
Calling the \code{eval} method on a query will convert it to executable
bytecode; this bytecode will be loaded and invoked by using Java reflection.
We produce a thunk that, when evaluated, will execute the
generated code.

@@

\section{Implementation: Expressing the interface in Scala}
\label{sec:intfScala}
\input{aosd13/pepm13/caseStudyImpl}

@@

\section{Language-Integrated Queries}
Microsoft's Language-Integrated Query technology (\LINQ)~\citep{Meijer:2006:LRO:1142473.1142552,Bierman:2007:LTF:1297027.1297063} is similar to our work in that it also
reifies queries on collections to enable analysis and optimization. Such queries can be executed against a variety of backends (such as SQL databases or in-memory objects), and adding new back-ends is supported. Its implementation uses \emph{expression trees}, a compiler-supported
implicit conversion between expressions and their reification as a syntax tree. There are various major differences, though.
First, the support for expression trees is hard-coded into the compiler. This means that the techniques are not applicable in languages 
that do not explicitly support expression trees. More importantly, the way expression trees are created in \LINQ\ is generic and fixed.
For instance, it is not possible to create different tree nodes for method calls that are relevant to an analysis (such as the \code{map} method) than for method calls that are irrelevant for the analysis (such as the \code{toString} method). For this reason, expression trees in \LINQ\ 
cannot be customized to the task at hand and contain too much low-level information. It is well-known that this makes it quite hard to
implement programs operating on expression trees~\citep{Eini11Pain}.

@@

\section{Query Optimization}
Query optimization on relational data is a long-standing issue in the database community, but there
are also many works on query optimization on objects~\citep{Fegaras00,Grust99PhD}.
Compared to these works, we have only implemented a few simple query optimizations, so there is potential
for further improvement of our work by incorporating more advanced optimizations.

@@

\section{Deep Embeddings in Scala}
\label{sec:rwdsl}
Technically, our implementation of \LoS\ is a deep embedding of a part of the Scala collections API~\citep{odersky2009fighting}.
Deep embeddings were pionereed by \citet{Leijen99DSEC} and \citet{elliott03compiling}. \pg{This makes no sense for the thesis.}
The technical
details of the embedding are not the main topic of this chapter; we are using some of the
Scala techniques presented by \citet{rompf2010lightweight} for using implicits and for
adding infix operators to a type. Similar to \citet{rompf2010lightweight}, we also 
use the Scala compiler on-the-fly. A plausible alternative backend for \LoS\ would
have been to use Delite~\citep{Rompf11BBlocks}, a framework for
building highly efficient DSLs in Scala.
Using this framework, in concurrent work, \citet{Rompf13} also optimize
collection queries; while their work allows for imperative programs, they do
not support embedding arbitrary libraries in an automated way. On the other
hand, they can reuse support for automatic parallelization and multiple platforms present in Delite.
\citet{Ackermann12} present Jet, which also optimizes collection queries but
targets MapReduce-style computations in a distributed environment.
Moreover, both works do not apply typical database optimizations such as
indexing or filter hoisting.

### 2017-01-18 aosd13/pepm13/caseStudyImpl.tex
aosd13: adjust relative paths
Revise sectioning
s/subsection/section for sections that became chapters of part I.

%\pg{Somewhere: Contrast a bit with LMS - we are more open, provide less compile-time safety, etc.; motivate differences. This section risks being boring, explain why it is not.}
\subsection{Representing expression trees}
\label{subsec:repexptrees}
In the previous section, we have seen that expressions that would have type \code{T} in a native Scala query are reified and have type \code{Exp[T]} in \LoS. The generic type \code{Exp[T]} is the base for our reification of Scala expression as expression trees, that is, as data structures in memory. We provide a subclass of \code{Exp[T]} for every different form of expression we want to reify. For example, in Fig.~\ref{fig:query} the expression \code{author.firstName + " " + author.lastName} must be reified even though it is not collection-related, for otherwise the optimizer could not see whether \code{author} is used. Knowing this is needed for instance to remove variables which are bound but not used. Hence, this expression is reified as \code{StringConcat(StringConcat(AuthorFirstName(author), Const(" ")), AuthorLastName(author))}. This example uses the constructors of the following subclasses of \code{Exp[T]} to create the expression tree.

@@

\subsection{Lifting first-order expressions}

@@

\subsection{Lifting higher-order expressions}
\label{sec:lifting-ho}
We have shown how to lift first-order expressions; however, the interface of collections also uses higher-order methods, that is, methods that accept functions as parameters, and we need to lift them as well to reify the complete collection DSEL. For instance, the \code{map} method applies a function to each element of a collection. In this subsection, we describe how we reify such expressions of function type.

\smartParagraph{Higher-order abstract syntax}
To represent functions, we have to represent the bound variables in the function bodies. For example, a reification of \code{str => str + "!"} needs to reify the variable \code{str} of type \code{String} in the body of the anonymous function. This reification should retain the information where \code{str} is bound. We achieve this by representing bound variables using \emph{higher-order abstract syntax} (HOAS)~\citep{Pfenning88hoas}, that is, we represent them by variables bound at the meta level. To continue the example, the above function is reified as \code{(str: Exp[String]) => str + "!"}. Note how the type of \code{str} in the body of this version is \code{Exp[String]}, because \code{str} is a reified variable now. Correspondingly, the expression \code{str + "!"} is lifted as described in the previous subsection.

@@

\smartParagraph{User-defined methods, revised}
We can now extend the lifting of signatures for methods or functions from the previous subsection to the general case, that is, the case of higher-order functions. We lift a method or function with signature
\begin{lstlisting}
def m[A1, $...$, An](a1: T1, $...$, an: Tn): R
\end{lstlisting}
to a method or function with the following signature.
\begin{lstlisting}
def m[A1, $...$, An](a1: $\ExpWrap\llbracket$T1$\rrbracket$, $...$, an: $\ExpWrap\llbracket$Tn$\rrbracket$): $\ExpWrap\llbracket$R$\rrbracket$
\end{lstlisting}
As before, the definition of the lifted method or function will return an expression node representing the call. If the original was a function, the lifted version is also defined as a function. If the original was a method on type \code{T}, then the lifted version is enriched onto \code{T}.

@@

\subsection{Lifting collections}
\label{subsec:collections}
In this subsection, we first explain how for-comprehensions are desugared to calls to library functions, allowing an external library to give them a different meaning. We summarize afterwards needed information about the subset of the Scala collection DSEL that we reify. Then we present how we perform this reification.
We finally present the reification of the running example (Fig.~\ref{fig:reifiedQuery}).

@@

%\pg{The following text is too abstract yet, but I need to write it down to clarify my ideas.}
%This subsection discusses how we lift operations on collections. We could apply the techniques in the previous sections directly, but three opposing design goals inform the design of our lifting:
%\begin{itemize}
%\item expressions in the Scala collections DSEL have an accurate static type~\citep{odersky2009fighting}, and we want to preserve that accuracy;
%\item we want our expression trees to be reasonably simple to transform; we want to have just a single type of node to represent \code{flatMap} operations, and not a different one for each kind of collection.
%\item we want to minimize the amount of code we need, and have minimal coupling to the collection library. For instance, the framework should support new implementation of dictionaries with as few changes as possible to the framework.
%\end{itemize}
%Furthermore, an additional non-functional constraint is that we need an interface which type inference understands.
%
%These three goals are opposing and we cannot reach all of them at the same time. In the end, we make the following compromises.
%\begin{itemize}
%\item We only support \code{Traversable[T]} and its subtypes. Some optimizations are unsafe for more generic collection types.
%For instance, we do not support iterators (whose interface relies on side effects), nor do we support \code{FilterMonadic[T, Repr]}.
%Additionally, this excludes support for parallel collections, which is left however for future work.
%\item We only support some operations yet (but the rest is left to future work).
%\item We do not support side-effectful interfaces.
%\end{itemize}

@@

%% PG: perhaps do say something more about indexes
%\subsection{Representing indexes: The indexBy operator}

%Indexes do not require special support; they are encoded through the
%\code{TraversableLike[T, Repr].indexBy[K](f: T => K): Map[K, Coll]} operator. A query of the form:
%
%\begin{lstlisting}
%coll withFilter (x => f(x) ==# y)
%\end{lstlisting}
%is semantically equivalent to a query which builds an index first with
%the result of applying \code{f} to elements of \code{coll}, and then
%selecting the entry for \code{y}:
%
%\begin{lstlisting}
%coll indexBy f apply y
%\end{lstlisting}
%What the optimizer does is (a variant of common subexpression
%elimination) between the resulting query and the keys of a map existing
%precomputed expressions. A repository contains a map from queries to
%their evaluated results (XXX: it should be incrementally maintained
%queries!).
%
%%XXX: show a query that we can help using indexing.
%
%%\subsection{Indexes for nested objects}
%
%Queries often traverse multiple objects, and more complex index
%organizations are needed.
%We reuse path indexes~\citep{Bertino89}.
%
\subsection{Interpretation}
\label{sec:interpretation}

@@

\subsection{Optimization}
%\pg{This example can now be optimized, the improvements took circa 3-4 hours.
%What's difficult is not at all going from Fig.~\ref{fig:titleFilterho1} to Fig.~\ref{fig:titleFilterho2}, it is just the 'inlining' steps, which seems to contain tons of intermediate normalization steps between other optimizations, fusion between filters and maps, and so on.
%Moreover, we need to perform inlining even when it causes work duplication, but the new optimizer seems to not behave worse on the evaluation data, even though it often replicates what it should not replicate; I tuned it to avoid inlining let expressions introduced explicitly by the user, but even before it didn't damage performance too much.
%The description of the optimizer should probably reflect the kind of needed steps, and cite also the paper on the Haskell inliner for future work.}
%

@@

%\pg{Select and integrate what needed.}
\smartParagraph{Manipulating functions}
%\subsection*{More about functions.}
%\tr{This material is left over from Sec. 4.3. It is partly needed in later sections and needs to be moved there.}
To be able to inspect a HOAS function body \code{funBody: Exp[S] => Exp[T]}, like \code{str => str + "!"}, we convert it to \emph{first-order} abstract syntax (FOAS), that is to an expression tree of type \code{Exp[T]}.
To this end, we introduce a representation of variables and a generator of fresh ones; since variable names are auto-generated, they are internally represented simply as integers instead of strings for efficiency.

@@

% \pg{For now comment out normalization by evaluation.}
%\subsection{Normalization by evaluation}
%\pg{Explain why we always convert to FOAS immediately and relate that to normalization by evaluation.}

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../aosd13-extended"
%%% End:

>>>>

%\pg{Somewhere: Contrast a bit with LMS - we are more open, provide less compile-time safety, etc.; motivate differences. This section risks being boring, explain why it is not.}
\section{Representing expression trees}
\label{subsec:repexptrees}
In the previous section, we have seen that expressions that would have type \code{T} in a native Scala query are reified and have type \code{Exp[T]} in \LoS. The generic type \code{Exp[T]} is the base for our reification of Scala expression as expression trees, that is, as data structures in memory. We provide a subclass of \code{Exp[T]} for every different form of expression we want to reify. For example, in Fig.~\ref{fig:query} the expression \code{author.firstName + " " + author.lastName} must be reified even though it is not collection-related, for otherwise the optimizer could not see whether \code{author} is used. Knowing this is needed for instance to remove variables which are bound but not used. Hence, this expression is reified as \code{StringConcat(StringConcat(AuthorFirstName(author), Const(" ")), AuthorLastName(author))}. This example uses the constructors of the following subclasses of \code{Exp[T]} to create the expression tree.

@@

\section{Lifting first-order expressions}

@@

\section{Lifting higher-order expressions}
\label{sec:lifting-ho}
We have shown how to lift first-order expressions; however, the interface of collections also uses higher-order methods, that is, methods that accept functions as parameters, and we need to lift them as well to reify the complete collection DSEL. For instance, the \code{map} method applies a function to each element of a collection. In this section, we describe how we reify such expressions of function type.

\smartParagraph{Higher-order abstract syntax}
To represent functions, we have to represent the bound variables in the function bodies. For example, a reification of \code{str => str + "!"} needs to reify the variable \code{str} of type \code{String} in the body of the anonymous function. This reification should retain the information where \code{str} is bound. We achieve this by representing bound variables using \emph{higher-order abstract syntax} (HOAS)~\citep{Pfenning88hoas}, that is, we represent them by variables bound at the meta level. To continue the example, the above function is reified as \code{(str: Exp[String]) => str + "!"}. Note how the type of \code{str} in the body of this version is \code{Exp[String]}, because \code{str} is a reified variable now. Correspondingly, the expression \code{str + "!"} is lifted as described in the previous section.

@@

\smartParagraph{User-defined methods, revised}
We can now extend the lifting of signatures for methods or functions from the previous section to the general case, that is, the case of higher-order functions. We lift a method or function with signature
\begin{lstlisting}
def m[A1, $...$, An](a1: T1, $...$, an: Tn): R
\end{lstlisting}
to a method or function with the following signature.
\begin{lstlisting}
def m[A1, $...$, An](a1: $\ExpWrap\llbracket$T1$\rrbracket$, $...$, an: $\ExpWrap\llbracket$Tn$\rrbracket$): $\ExpWrap\llbracket$R$\rrbracket$
\end{lstlisting}
As before, the definition of the lifted method or function will return an expression node representing the call. If the original was a function, the lifted version is also defined as a function. If the original was a method on type \code{T}, then the lifted version is enriched onto \code{T}.

@@

\section{Lifting collections}
\label{subsec:collections}
In this section, we first explain how for-comprehensions are desugared to calls to library functions, allowing an external library to give them a different meaning. We summarize afterwards needed information about the subset of the Scala collection DSEL that we reify. Then we present how we perform this reification.
We finally present the reification of the running example (Fig.~\ref{fig:reifiedQuery}).

@@

%\pg{The following text is too abstract yet, but I need to write it down to clarify my ideas.}
%This section discusses how we lift operations on collections. We could apply the techniques in the previous sections directly, but three opposing design goals inform the design of our lifting:
%\begin{itemize}
%\item expressions in the Scala collections DSEL have an accurate static type~\citep{odersky2009fighting}, and we want to preserve that accuracy;
%\item we want our expression trees to be reasonably simple to transform; we want to have just a single type of node to represent \code{flatMap} operations, and not a different one for each kind of collection.
%\item we want to minimize the amount of code we need, and have minimal coupling to the collection library. For instance, the framework should support new implementation of dictionaries with as few changes as possible to the framework.
%\end{itemize}
%Furthermore, an additional non-functional constraint is that we need an interface which type inference understands.
%
%These three goals are opposing and we cannot reach all of them at the same time. In the end, we make the following compromises.
%\begin{itemize}
%\item We only support \code{Traversable[T]} and its subtypes. Some optimizations are unsafe for more generic collection types.
%For instance, we do not support iterators (whose interface relies on side effects), nor do we support \code{FilterMonadic[T, Repr]}.
%Additionally, this excludes support for parallel collections, which is left however for future work.
%\item We only support some operations yet (but the rest is left to future work).
%\item We do not support side-effectful interfaces.
%\end{itemize}

@@

%% PG: perhaps do say something more about indexes
%\section{Representing indexes: The indexBy operator}

%Indexes do not require special support; they are encoded through the
%\code{TraversableLike[T, Repr].indexBy[K](f: T => K): Map[K, Coll]} operator. A query of the form:
%
%\begin{lstlisting}
%coll withFilter (x => f(x) ==# y)
%\end{lstlisting}
%is semantically equivalent to a query which builds an index first with
%the result of applying \code{f} to elements of \code{coll}, and then
%selecting the entry for \code{y}:
%
%\begin{lstlisting}
%coll indexBy f apply y
%\end{lstlisting}
%What the optimizer does is (a variant of common subexpression
%elimination) between the resulting query and the keys of a map existing
%precomputed expressions. A repository contains a map from queries to
%their evaluated results (XXX: it should be incrementally maintained
%queries!).
%
%%XXX: show a query that we can help using indexing.
%
%%\section{Indexes for nested objects}
%
%Queries often traverse multiple objects, and more complex index
%organizations are needed.
%We reuse path indexes~\citep{Bertino89}.
%
\section{Interpretation}
\label{sec:interpretation}

@@

\section{Optimization}
%\pg{This example can now be optimized, the improvements took circa 3-4 hours.
%What's difficult is not at all going from Fig.~\ref{fig:titleFilterho1} to Fig.~\ref{fig:titleFilterho2}, it is just the 'inlining' steps, which seems to contain tons of intermediate normalization steps between other optimizations, fusion between filters and maps, and so on.
%Moreover, we need to perform inlining even when it causes work duplication, but the new optimizer seems to not behave worse on the evaluation data, even though it often replicates what it should not replicate; I tuned it to avoid inlining let expressions introduced explicitly by the user, but even before it didn't damage performance too much.
%The description of the optimizer should probably reflect the kind of needed steps, and cite also the paper on the Haskell inliner for future work.}
%

@@

%\pg{Select and integrate what needed.}
\smartParagraph{Manipulating functions}
%\section*{More about functions.}
%\tr{This material is left over from Sec. 4.3. It is partly needed in later sections and needs to be moved there.}
To be able to inspect a HOAS function body \code{funBody: Exp[S] => Exp[T]}, like \code{str => str + "!"}, we convert it to \emph{first-order} abstract syntax (FOAS), that is to an expression tree of type \code{Exp[T]}.
To this end, we introduce a representation of variables and a generator of fresh ones; since variable names are auto-generated, they are internally represented simply as integers instead of strings for efficiency.

@@

% \pg{For now comment out normalization by evaluation.}
%\section{Normalization by evaluation}
%\pg{Explain why we always convert to FOAS immediately and relate that to normalization by evaluation.}

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../../main"
%%% End:
